signalHsmm - a novel semi-Markov model of eukaryotic signal peptides

### 2015-06-16 paper.tex
adding references to paper that use hsmm in bioinformatics

\subsection*{Hidden semi-Markov model}

Let us first focus on Hidden Markov Model (HMM). 
Suppose you have a sequence of observations e.g. amino acids. What is of interest is understanding underlying cause of their occurrence. 
HMM aims to answer that by assuming a specific yet flexible structure of the problem.
HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
These are "the cause" for the observations. At every step, hidden state might change according to transition matrix
$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application hidden states are signal peptide regions.
Second process $E_{t=1}^T$ is observation process, defined on the set of possible observations $\{O_1, \dots, O_m\}$. 
Observations are assumed to occur independently. Their distribution depends only on hidden state which emits it. 
Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
In our case, observations are (degenerated) amino acids.
Main goal in signalHsmm is to find, for a given peptide, most probable regions boundaries. This is achieved with
Viterbi algorithm.
For a good reference on HMM see~\citep{1989rabinera}.

@@

\node[block,text width=3.5em, bottom color=green!50!red!90] at (8+2,0) (block3) {\footnotesize $k^{th}$ hidden state};
\node[blockComp] at (6.5+2,-2)  (komp5) {\footnotesize \pgfmathparse{\names[4]}\pgfmathresult};
\node[blockComp] at (8+2,-2)  (komp6) {\footnotesize \pgfmathparse{\names[5]}\pgfmathresult};
\node[blockComp] at (9.5+2,-2)  (komp7) {\footnotesize \pgfmathparse{\names[6]}\pgfmathresult};
\draw[line,->]  (block3) --  (komp5);
\draw[line,->]  (block3) --  (komp6);
\draw[line,->]  (block3) --  (komp7);

>>>>

\subsection*{Hidden semi-Markov model}
Hidden semi-Markov model (HSMM) is an extension of popular method  hidden Markov model (HMM).
In many fields it proved to yield better accuracy of bioinformatics algorithms.
Gene recognition~\citep{Pachter02applicationsof}, protein secondary structure prediction  \citep{16571137}
or automatic detection of heartbeats~\citep{7043102}.
Let us first focus on hidden Markov model (HMM). 
Suppose you have a sequence of observations e.g. amino acids. What is of interest is understanding underlying cause of their occurrence. 
HMM aims to answer that by assuming a specific yet flexible structure of the problem.
HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
These are "the cause" for the observations. At every step, hidden state might change according to transition matrix
$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application hidden states are signal peptide regions.
Second process $E_{t=1}^T$ is observation process, defined on the set of possible observations $\{O_1, \dots, O_m\}$. 
Observations are assumed to occur independently. Their distribution depends only on hidden state which emits it. 
Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
In our case, observations are (degenerated) amino acids.
Main goal in signalHsmm is to find, for a given peptide, most probable regions boundaries. This is achieved with
Viterbi algorithm.
For a good reference on HMM see~\citep{1989rabinera}.

@@

\node[block,text width=3em, bottom color=green!50!red!90] at (8+2,0) (block3) {\footnotesize $k^{th}$ hidden state};
\node[blockComp] at (6.5+2,-2)  (komp5) {\footnotesize \pgfmathparse{\names[4]}\pgfmathresult};
\node[blockComp] at (8+2,-2)  (komp6) {\footnotesize \pgfmathparse{\names[5]}\pgfmathresult};
\node[blockComp] at (9.5+2,-2)  (komp7) {\footnotesize \pgfmathparse{\names[6]}\pgfmathresult};
\draw[line,->]  (block3) --  (komp5);
\draw[line,->]  (block3) --  (komp6);
\draw[line,->]  (block3) --  (komp7);
